---
title: "RepResCour2"
author: "CP"
date: "2023-08-02"
output: html_document
---

#Synopsis
This study investigates the severe weather events in the US based on time series data from the National 
Oceanic and Atmospheric Administration (NOAA). The NOAA storm database not only tracks a lot of events 
across all states in the US, but also provides information on injuries and property damages.
For a detailled description of the NOAA storm database see: https://www.ncdc.noaa.gov/stormevents/
The present analysis makes use of that information and clearly shows that tornados have the most 
harmful impact on people’s health, whereas floods are most expensive from an economic view.

#Data Processing
Load packages
```{r}
library(dplyr)
library(lubridate)
library(ggplot2)
```

Load data
```{r}
if (!file.exists("StormData.csv.bz2")) {
  fileUrl<-"https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
  download.file(fileUrl, destfile="StormData.csv.bz2")
  # Exit if the file is not available
  if (!file.exists("StormData.csv.bz2")) {
    stop("Can't locate file 'StormData.csv.bz2'!")
  }
}
```

Load the dataset
```{r}
stormDataRaw <- read.csv("StormData.csv.bz2")
```

Show the structure of the dataset
```{r}
str(stormDataRaw)
```

Data subset
```{r}
stormData <- select(stormDataRaw, BGN_DATE, EVTYPE, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP, FATALITIES, INJURIES)
```

Format the BGN_DATE variable as a date
```{r}
stormData$BGN_DATE <- as.Date(stormData$BGN_DATE, "%m/%d/%Y")
stormData$YEAR <- year(stormData$BGN_DATE)
```

Only use events since 1996
```{r}
stormData <- filter(stormData, YEAR >= 1996)
```

Only use events with either health impact or economic damage
```{r}
stormData <- filter(stormData, PROPDMG > 0 | CROPDMG > 0 | FATALITIES > 0 | INJURIES > 0)
```

Adjustment economical damages
```{r}
table(stormData$PROPDMGEXP)
table(stormData$CROPDMGEXP)

stormData$PROPDMGEXP <- toupper(stormData$PROPDMGEXP)
stormData$CROPDMGEXP <- toupper(stormData$CROPDMGEXP)

stormData$CROPDMGFACTOR[(stormData$CROPDMGEXP == "")] <- 10^0
stormData$CROPDMGFACTOR[(stormData$CROPDMGEXP == "?")] <- 10^0
stormData$CROPDMGFACTOR[(stormData$CROPDMGEXP == "0")] <- 10^0
stormData$CROPDMGFACTOR[(stormData$CROPDMGEXP == "2")] <- 10^2
stormData$CROPDMGFACTOR[(stormData$CROPDMGEXP == "K")] <- 10^3
stormData$CROPDMGFACTOR[(stormData$CROPDMGEXP == "M")] <- 10^6
stormData$CROPDMGFACTOR[(stormData$CROPDMGEXP == "B")] <- 10^9

stormData$PROPDMGFACTOR[(stormData$PROPDMGEXP == "")] <- 10^0
stormData$PROPDMGFACTOR[(stormData$PROPDMGEXP == "-")] <- 10^0
stormData$PROPDMGFACTOR[(stormData$PROPDMGEXP == "?")] <- 10^0
stormData$PROPDMGFACTOR[(stormData$PROPDMGEXP == "+")] <- 10^0
stormData$PROPDMGFACTOR[(stormData$PROPDMGEXP == "0")] <- 10^0
stormData$PROPDMGFACTOR[(stormData$PROPDMGEXP == "1")] <- 10^1
stormData$PROPDMGFACTOR[(stormData$PROPDMGEXP == "2")] <- 10^2
stormData$PROPDMGFACTOR[(stormData$PROPDMGEXP == "3")] <- 10^3
stormData$PROPDMGFACTOR[(stormData$PROPDMGEXP == "4")] <- 10^4
stormData$PROPDMGFACTOR[(stormData$PROPDMGEXP == "5")] <- 10^5
stormData$PROPDMGFACTOR[(stormData$PROPDMGEXP == "6")] <- 10^6
stormData$PROPDMGFACTOR[(stormData$PROPDMGEXP == "7")] <- 10^7
stormData$PROPDMGFACTOR[(stormData$PROPDMGEXP == "8")] <- 10^8
stormData$PROPDMGFACTOR[(stormData$PROPDMGEXP == "H")] <- 10^2
stormData$PROPDMGFACTOR[(stormData$PROPDMGEXP == "K")] <- 10^3
stormData$PROPDMGFACTOR[(stormData$PROPDMGEXP == "M")] <- 10^6
stormData$PROPDMGFACTOR[(stormData$PROPDMGEXP == "B")] <- 10^9
```

Economic impact
```{r}
stormData <- mutate(stormData, HEALTHIMP = FATALITIES + INJURIES)
stormData <- mutate(stormData, ECONOMICCOST = PROPDMG * PROPDMGFACTOR + CROPDMG * CROPDMGFACTOR)
```

Event type dimension
```{r}
stormData$EVTYPE <- toupper(stormData$EVTYPE)
dim(data.frame(table(stormData$EVTYPE)))
```

This section shows repetition of events due to inconsistent event names
```{r}
evtypeUnique <- unique(stormData$EVTYPE)
evtypeUnique[grep("THUND", evtypeUnique)]
```

Clean-up event type
```{r}
healthImpact <- with(stormData, aggregate(HEALTHIMP ~ EVTYPE, FUN = sum))
subset(healthImpact, HEALTHIMP > quantile(HEALTHIMP, prob = 0.95))
```
```{r}
stormData$EVTYPE[(stormData$EVTYPE == "TSTM WIND")] <- "THUNDERSTORM WIND"
stormData$EVTYPE[(stormData$EVTYPE == "HURRICANE/TYPHOON")] <- "HURRICANE (TYPHOON)"
```
```{r}
economicCost <- with(stormData, aggregate(ECONOMICCOST ~ EVTYPE, FUN = sum))
subset(economicCost, ECONOMICCOST > quantile(ECONOMICCOST, prob = 0.95))
```

Two types of events in the 95% quantile
```{r}
stormData$EVTYPE[(stormData$EVTYPE == "HURRICANE")] <- "HURRICANE (TYPHOON)"
stormData$EVTYPE[(stormData$EVTYPE == "STORM SURGE")] <- "STORM SURGE/TIDE"
```

#Results

Aggregate cleaned data frame per EVTYPE for fatalities and injuries
```{r}
healthImpact <- stormData %>% 
                group_by(EVTYPE) %>% 
                summarise(HEALTHIMP = sum(HEALTHIMP)) %>% 
                arrange(desc(HEALTHIMP))
#healthImpact[1:10,]
g1 <- ggplot(healthImpact[1:10,], aes(x=reorder(EVTYPE, -HEALTHIMP),y=HEALTHIMP,color=EVTYPE)) + 
      geom_bar(stat="identity", fill="white") + 
      theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
      xlab("Event") + ylab("Number of fatalities and injuries") +
      theme(legend.position="none") +
      ggtitle("Fatalities and injuries in the US caused by severe weather events")
g1
```

The barchart shows that Tornados are the most harmful weather events for people’s health.

Aggregate cleaned data frame per EVTYPE for economic cost
```{r}
economicCost <- stormData %>% 
                group_by(EVTYPE) %>% 
                summarise(ECONOMICCOST = sum(ECONOMICCOST)) %>% 
                arrange(desc(ECONOMICCOST))
#economicCost[1:10,]
g2 <- ggplot(economicCost[1:10,], aes(x=reorder(EVTYPE, -ECONOMICCOST),y=ECONOMICCOST,color=EVTYPE)) + 
      geom_bar(stat="identity", fill="white") + 
      theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
      xlab("Event") + ylab("Economic cost in USD") +
      theme(legend.position="none") +
      ggtitle("Economic cost in the US caused by severe weather events")
g2
```

